---
title: "Plotting"
---

# Plotting data

Plotting of data in modern R is handled by 
[ggplot2](https://ggplot2.tidyverse.org), which is installed as part of the `tidyverse`.
There is an [excellent cheat sheet](https://github.com/rstudio/cheatsheets/blob/master/data-visualization-2.1.pdf)
and a great free online book,
[ggplot2: Elegant graphics for data analysis, by Hadley Wickham, Danielle Navarro, and Thomas Lin Pedersen](https://ggplot2-book.org).

We cannot cover all of [ggplot2](https://ggplot2.tidyverse.org) in this workshop, 
so you are strongly encouraged to read the book and cheat sheet. For today,
we will give a quick overview of how to create some simple graphs.

First we load the `tidyverse` in the same way as previously;

```{r}
library(tidyverse)
```

Next we will load some data to plot. We will use the climate data from the previous
section (available from [cetml1659on.txt](https://raw.githubusercontent.com/Bristol-Training/intro-data-analysis-r/refs/heads/main/data/cetml1659on.txt)).

Plotting works best with tidy data, so we will load and tidy the data as in the previous
section;

```{r}
temperature <- read_table(
    "https://raw.githubusercontent.com/Bristol-Training/intro-data-analysis-r/refs/heads/main/data/cetml1659on.txt",
    skip=6,
    na=c("-99.99", "-99.9"),
    col_types=cols("DATE"=col_integer())
)

month_levels <- c("JAN", "FEB", "MAR", "APR", "MAY", "JUN",
                  "JUL", "AUG", "SEP", "OCT", "NOV", "DEC")

historical_temperature <- temperature %>%
    select(-YEAR) %>%
    pivot_longer(c("JAN", "FEB", "MAR", "APR", "MAY", "JUN",
                   "JUL", "AUG", "SEP", "OCT", "NOV", "DEC"),
                 names_to="month",
                 values_to="temperature") %>%
    rename(year=DATE) %>%
    mutate(month=factor(month, month_levels))
```

Next, we will use `ggplot` to draw a graph (we will explain how 
this works after drawing).
You should see a graph similar to this:

```{r}
ggplot(historical_temperature, aes(x = year, y = temperature)) + geom_point()
```


This command has drawn a scatter plot of the data contained in the tibble
`historical_temperature`, putting the `year` column on the x-axis, and
the `temperature` column on the y-axis.

`ggplot` is written to follow a specific "grammar of visualisation". 
There are three key components;

1. data,
2. A set of *aesthetic mappings* between variables in the data and 
   visual properties of the graph, and
3. one or more *layers* that describe how to render each observation.

These are entered via the general form

```
ggplot( data, aesthetic ) + layer1 + layer2 + layer3...
```

The `data` is a tibble containing tidy data with one observation per row,
and one variable per column.

The `aesthetic` is a mapping specified via the `aes` function, which
maps the variables to axes, colours or other graphical properties.

The layers (`layer1`, `layer2` etc) are specific renderings of the data,
e.g. `geom_point()` will draw points (scatter plot), `geom_line()` 
will draw lines (line graph) etc.

## Plotting the analysis

You can combine analysis with plotting, e.g. here we plot the average
yearly temperature as a line graph.

```{r}
ggplot(historical_temperature %>%
           group_by(year) %>%
           summarise(temperature=mean(temperature)), 
       aes(x = year, y = temperature)) +
    geom_line() + 
    geom_smooth()
```

Here we've added two layers; `geom_line()` to draw a line graph, 
and `geom_smooth()` to add a fitted line with errors.


We could do more. For example, here we add a new variable (column)
to the data that is the decade in which the observation was taken;

```{r}
historical_temperature["decade"] <- (historical_temperature["year"] %/% 10) * 10
```

(`%/%` means "integer division", so `1655 %/% 10` equals `165`, which
becomes `1650` when multiplied by `10`)

This enables us to draw a line graph of the average temperature 
each decade;

```{r}
ggplot(historical_temperature %>%
           group_by(decade) %>%
           summarise(temperature=mean(temperature)), 
       aes(x = decade, y = temperature)) + 
       geom_line()
```

A line chart is not a good choice for this plot, as it doesn't
show the underlying statistics of the average. Instead, a
box-and-whisker or violin plot would be better. To use this,
we need to specify the grouping in the aesthetic, e.g.

```{r}
ggplot(historical_temperature, 
       aes(x = decade, y = temperature, group=decade)) + 
       geom_violin()
```

Finally, the global aesthetic set in `ggplot` can be overridden by 
setting it in the layers themselves. For example, here we overlay
a smooth line over the violin plot;

```{r}
ggplot(historical_temperature, 
       aes(x=year, y = temperature)) + 
       geom_violin(aes(group=decade)) + 
       geom_smooth()
```

Note how the group aesthetic has to be set only for the violin plot.

You can save your plots to a file using the 
[ggsave](https://ggplot2.tidyverse.org/reference/ggsave.html) function.
This will save the last plot drawn to a file, with filename, size,
format etc. all controlled via arguments to this function, e.g.

```{.r}
ggsave("violin.pdf", device="pdf", dpi="print")
```

would save the plot to a file called `violin.pdf`, in PDF format,
using a resolution (dpi) that is suitable for printing.




::: {#exerciseN .callout-note icon=false title='Exercise'}
Create a graph that shows the average maximum temperature by month. Draw this as a line graph. Note that you may need to add `group="decade"` to the aesthetic so that the line layer can join together the points.

Create a graph that shows the change in average temperature in December per decade. Draw this as a scatter plot, with a smooth trend line added.

Create a graph that shows the change in average temperature by century. Draw this as a line graph with a smooth trend line added. Note that you may need to adjust the [span](https://ggplot2.tidyverse.org/reference/geom_smooth.html) value if the number of data points is too few to draw a very smooth line.
:::

::: {#answerN .callout-caution icon=false title='Answer' collapse="true"}
First load the `tidyverse`, then read in all of the data and tidy it up...

```{r}
library(tidyverse)

temperature <- read_table(
    "https://raw.githubusercontent.com/Bristol-Training/intro-data-analysis-r/refs/heads/main/data/cetml1659on.txt",
    skip=6,
    na=c("-99.99", "-99.9"),
    col_types=cols("DATE"=col_integer())
)

month_levels <- c("JAN", "FEB", "MAR", "APR", "MAY", "JUN",
                  "JUL", "AUG", "SEP", "OCT", "NOV", "DEC")

historical_temperature <- temperature %>%
    select(-YEAR) %>%
    pivot_longer(c("JAN", "FEB", "MAR", "APR", "MAY", "JUN",
                   "JUL", "AUG", "SEP", "OCT", "NOV", "DEC"),
                 names_to="month",
                 values_to="temperature") %>%
    rename(year=DATE) %>%
    mutate(month=factor(month, month_levels))
```

Next, add in the "decade" and "century" variables (columns) as we will
be using them later...

```{r}
historical_temperature["decade"] <- (historical_temperature["year"] %/% 10) * 10
historical_temperature["century"] <- (historical_temperature["year"] %/% 100) * 100
historical_temperature
```

Graph that shows the average maximum temperature per month...

```{r}
ggplot(historical_temperature %>%
         group_by(month) %>%
         summarise(max_temp=max(temperature, na.rm=TRUE), .groups="drop"),
       aes(x=month, y=max_temp, group="month")) + 
  geom_line()
```

Change in average temperature in December per decade.

```{r}
ggplot( historical_temperature %>%
          filter(month=="DEC") %>%
          group_by(decade) %>%
          summarise(average=mean(temperature, na.rm=TRUE), .groups="drop"),
        aes(x=decade, y=average)
      ) + 
  geom_point() +
  geom_smooth()
```

Change in average temperature per century, plus a smooth trend line

```{r}
ggplot( historical_temperature %>%
          group_by(century) %>%
          summarise(average=mean(temperature, na.rm=TRUE), .groups="drop"),
        aes(x=century, y=average)) + 
  geom_point() + 
  geom_smooth(span=2.0)
```
:::


